---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

# Executive Summary

This report summarizes the relationship between MPG and transmission types in the mtcars dataset. The result of the statistical analysis using linear regression models is that a switch from automatic to manual transmission should lead to an **increase in miles per gallon**.  


## Data: mtcars

Change format of variable 'am' to factor and review structure of 'mtcars'. 

```{r}
library(data.table)
library(tinytex)
# tinytex::install_tinytex()
attach(mtcars)
mtcars <- as.data.table(mtcars)
levels(mtcars$am)
mtcars$am <- factor(mtcars$am, labels = c("auto", "man"))
str(mtcars)
```
Transmission is represented by a factor variable 'am' where 0 stands for automatic and 1 represents manual.  


\newpage
## Scatterplots

### Scatterplot mpg ~ disp

```{r}
plot(disp, mpg)
abline(lm(mpg ~ disp))
```
\newpage
### Scatterplot mpg ~ wt
```{r}
plot(wt, mpg)
abline(lm(mpg ~ wt))
```

\newpage
## Fit linear model with reasonable regressors

Fit a model including all regressors that seem reasonable, i.e. are expected to have an effect on 'miles per gallon':  

```{r fit model with reasonable regressors}
fit_reas <- lm(mpg ~ disp + wt + am - 1)
summary(fit_reas)
```
Switching from automatic to manual increases miles per gallons.  

Extremely low **p-value** shows that result is **statistically significant**.  

\newpage
### Check Residuals' Plot

```{r}
par(mfrow = c(2,2))
plot(fit_reas)
```

\newpage
# Only 'am' as Regressor

```{r only transmission as regressor}
fit_am <- lm(mpg ~ am)
summary(fit_am)
```
Result statistically significant as the p-value is far below 0.05.  
Going from 'automatic' to 'manual' the **miles per gallons increases**.  

\newpage
### Check Residuals' Plot
```{r}
par(mfrow = c(2,2))
plot(fit_reas)
```

# t-Test

```{r}
t.test(mpg ~ am, mtcars)
```
The difference between the means is statistically significant considering the low p-value. 
